---
title: "DDS Project 2"
author: "Mohammad Al-Rousan"
date: "2023-04-12"
output: html_document
---

## Entering the Data and Correct librarys
```{r}
#Import the libraries
library(XML)
library(dplyr)
library(tidyr)
library(stringi)
library(rvest) 
library(ggplot2)
library(RCurl)
library(class)
library(caret)
library(e1071)
library(stringr)
library(naniar)
library(rmarkdown)
library(readxl)
library(tidyverse)
library(aws.s3)

#Import the information from AWS
Sys.setenv("AWS_ACCESS_KEY_ID" = "AKIAUWXEX5B6B23VEXHZ",
           "AWS_SECRET_ACCESS_KEY" = "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",
           "AWS_DEFAULT_REGION" = "us-east-2")
aws.s3::bucketlist()

aws.s3::get_bucket("ddsproject1")

#Importing and reading the data 
employeeClassEx = s3read_using(FUN = read.csv,
                           bucket = "ddsproject1",
                           object = "Case2PredictionsClassifyEXAMPLE.csv")

head(employeeClassEx)

employeePredEx = s3read_using(FUN = read.csv,
                               bucket = "ddsproject1",
                               object = "Case2PredictionsRegressEXAMPLE.csv")

head(employeePredEx)

employeePredEx = s3read_using(FUN = read.csv,
                              bucket = "ddsproject1",
                              object = "Case2PredictionsRegressEXAMPLE.csv")

employee = s3read_using(FUN = read.csv,
                              bucket = "ddsproject1",
                              object = "CaseStudy2-data.csv")
employeeAtt = s3read_using(FUN = read.csv,
                        bucket = "ddsproject1",
                        object = "CaseStudy2CompSet No Attrition.csv")

employeeSal = s3read_using(FUN = read.csv,
                          bucket = "ddsproject1",
                          object = "CaseStudy2CompSet No Salary.csv")

employeeAtt <- na.omit(employeeAtt)
is.na(employeeAtt)
View(employee)
```
## Data observations

```{r Obserations}
employee %>% group_by(Attrition) %>% count(Attrition)
#Currently, the company has a 16% attrition rate, meaning employees will leave 16%, but based on what factors? 

#I would like to take a deep dive into the dynamics of the company, who they hire, what educatoinal fields they, their average salaries, a numerous factors that will be further explained in detal 

employee%>% count(EducationField)
employee %>% count(JobRole)
#This company made hires in educational fields including Human resources, life sciences, marketing, technical degrees, and other. 
#These fields hire for job roles including HealthCare Representative, Human Resources, Laboratory Technician, Manager, Manufacturing Director, Research Director, Research Scientist, and Sales Executive. 

employee %>% summarise(mean(MonthlyIncome))
employee %>% summarise(mean(MonthlyRate))
#There is an aver Monthly Income of $6390.26 and a monthly rate of $14,325.62

employee %>% summarise(mean(JobSatisfaction))
#The company has an aver satisfactor level of 2.7, which is about above average meaning the majority of the employees are enjoying their jobs 

#These are the basic ideals people know about a company, what kind of people do they hire, how much they pay, and are they generally happy.
```

## Important factors and their reason

You can also embed plots, for example:

```{r pressure, echo=FALSE}
#First, I believe observing the montary consumption is a reason as to why those are leaving the jobs
employee %>% ggplot(aes(x = Attrition, y = MonthlyIncome, fill = Attrition)) +
  ylab("Monthly Income") +
  xlab("Attrition") +
  geom_boxplot() + 
  ggtitle("Attrition and Monthly Income")

employee %>% ggplot(aes(x = Attrition, y = MonthlyRate, fill = Attrition)) +
  ylab("Monthly Income") +
  xlab("Attrition") +
  geom_boxplot() + 
  ggtitle("Attrition and Monthly Rate")

employee %>% group_by(Attrition) %>% summarise(compareincomes = mean(MonthlyIncome))

#Based on the graphs, it would be easy to say that attrtion is occuring due to monetary consumption, but that would be a copout, the best stategy is to locate factors in their salary as to why they are leaving. 

#The first variable to observe is business travel
employee %>% count(BusinessTravel)
#The statistical breakdown is 10% of employees do not travel, 71% of employee's rarely travel, and 18% of employees travel frequently
employee %>% group_by(Attrition) %>% count(BusinessTravel)
#The statistical summary contents that 13.25% of employees who do not travel leave, 17.93% of those who rarely travel leave, but 28.45% of those who travel frequently leave, meaning those who travel frequently for their work will have a higher chance in leaving and we will create a graph showing how does it compare to monetary value. 
employee %>% group_by(BusinessTravel) %>% summarise(CompareIncomes = mean(MonthlyIncome))
# No Travel $14,500.78
#Travel Rarely $14,418.82
#Travel Frequently $13856.86	

employee %>% ggplot(aes(x = BusinessTravel, y = MonthlyIncome, fill = Attrition)) +
  xlab("Business Travel ") +
  ylab("Monthly Income") +
  geom_boxplot() +
  ggtitle("Attrition and Business Travel")

employee %>% ggplot(aes(x = BusinessTravel, y = MonthlyRate, fill = Attrition)) +
  xlab("Business Travel ") +
  ylab("Monthly Rate") +
  geom_boxplot() +
  ggtitle("Attrition and Business Travel")

#So, now we see that Business travel plays a very distinct role in one's salary and their, so now we provide examples on jobs and department that have the most travel leaving will have a high rate of attrition

employee %>% group_by(MaritalStatus) %>% count(BusinessTravel)
#Based on this statisitical summary, martial status plays an important role, and those who are single travel the most frequently which is 23.04%

employee %>% ggplot(aes(x = Attrition, fill = MaritalStatus)) + 
  geom_bar() + 
  facet_wrap(~BusinessTravel) +
  ylab("Count") +
  xlab("Attrition") +
  ggtitle ("Attrition and Marital Status")

employee %>% ggplot(aes(x = MonthlyRate, y = MaritalStatus, color = BusinessTravel)) +
  geom_point(position = "jitter") + 
  facet_wrap(~Attrition) +
  ylab("Marital Status") +
  xlab("Monthy Rate") +
  ggtitle("Attrtion and Travel")

employee %>% group_by(BusinessTravel) %>% summarise(mean(StockOptionLevel))
employee %>% group_by(StockOptionLevel) %>% summarise(mean(MonthlyIncome))

employee %>% ggplot(aes(x = StockOptionLevel, fill = Attrition)) + 
  geom_histogram() +
  facet_grid(~MaritalStatus) +
  xlab("Marital Status") +
  ylab("Count") +
  ggtitle("Attrition vs. Martial Status and Stock Option")


employee %>% ggplot(aes(x = StockOptionLevel, y = MonthlyIncome, color = Attrition)) + 
  geom_point(position = "jitter") + 
  facet_grid(~MaritalStatus) +
  xlab("Marital Status") +
  ylab("Monthly Income") +
  ggtitle("Attrition vs. Martial Status and Monthly Income")
  

employee %>% ggplot(aes(x = MonthlyIncome, y = StockOptionLevel, color = Attrition)) + 
  geom_point(position = "jitter") + 
  facet_grid(~BusinessTravel) +
  xlab("Monthly Income") +
  ylab("Stock Option") +
  ggtitle("Attrition vs. Stock Option and Monthly Income")
  

employee %>% group_by(OverTime) %>% summarise(mean(MonthlyIncome))
employee %>% ggplot(aes(x = OverTime, fill = Attrition)) + 
  geom_bar() + 
  facet_wrap(~MaritalStatus) +
  xlab("Monthly Income") +
  ylab("Stock Option") +
  ggtitle("Attrition vs. Stock Option and Monthly Income")
  
employee %>% ggplot(aes(x = OverTime, y = MonthlyIncome, color = Attrition)) +
  geom_point(position = "jitter") + 
  facet_wrap(~MaritalStatus) +
  xlab("Over Time") +
  ylab("Monthly Income") +
  ggtitle("Attrition vs. Over Time and Monthly Income")

#This is interesting to see this graph, because clearly the single group earns the least, travels the most, and have the highest attrition rate. 


```
```

#Creating a Regression Model
```{r}
predictionFit = lm(MonthlyIncome~JobLevel+JobRole+TotalWorkingYears+DistanceFromHome+
                     Age+PerformanceRating+YearsAtCompany+YearsSinceLastPromotion+OverTime
                    + MaritalStatus + BusinessTravel + StockOptionLevel 
                   + WorkLifeBalance + YearsAtCompany + Attrition,data=employee)

summary(predictionFit)
plot(predictionFit)

MonthlyEarning = predict(predictionFit,newdata=employeeSal)
miPred = cbind(employeeSal,MonthlyEarning)
miPred = subset(miPred,select=c(ID,MonthlyEarning))

cm(predictionFit)


employee$lmPred = predictionFit$fitted.values
ggplot(employee,aes(x=lmPred,y=MonthlyIncome))+
  geom_point()+
  geom_smooth(method='lm')+
  ggtitle('Linear Regression for Monthly Income')+
  xlab('Predicted Values')+
  ylab('Monthly Income')
```

#Creating A Naive Bayes Model
```{r}
employeeNB = employee
employeeNB$Attrition = as.factor(employee$Attrition)

splitPerc = 0.7

iterations = 100
employeeAcc = matrix(nrow = iterations)
employeeSen = matrix(nrow = iterations)
employeeSpec = matrix(nrow = iterations)

for(j in 1:iterations) {
  trainIndices = sample(seq(1: length(employee$ID)), round(splitPerc * length(employee$ID)))
  employeeTrain = employee[trainIndices, ]
  employeeTest = employee[-trainIndices, ]
  
  payNB <- naiveBayes(Attrition~ MonthlyIncome +JobLevel+JobRole+TotalWorkingYears+DistanceFromHome+
                        Age+PerformanceRating+YearsAtCompany+YearsSinceLastPromotion+OverTime
                      + MaritalStatus + BusinessTravel + StockOptionLevel 
                      + WorkLifeBalance + YearsAtCompany, data = employeeTrain, laplace = 1)
  
  table(predict(payNB, employeeTest), test$Attrition)
  
  CM = confusionMatrix(table(predict(payNB, employeeTest), employeeTest$Attrition))
  employeeAcc[j] = CM$overall[1]
  employeeSen[j] = CM$byClass[1]
  employeeSpec[j] = CM$byClass[2]
}

employeeAcc = colMeans(employeeAcc)
employeeAcc

employeeSen = colMeans(employeeSen)
employeeSen

employeeSpec = colMeans(employeeSpec)
employeeSpec

#The model gave a 87% sensitivity and a 60% specificity. 
```
#Finding the Columns
```{r}
grep("MonthlyIncome", colnames(employeeAtt))
grep("JobLevel", colnames(employeeAtt))
grep("TotalWorkingYears", colnames(employeeAtt))
grep("DistanceFromHome", colnames(employeeAtt))
grep("Age", colnames(employeeAtt))
grep("PerformanceRating", colnames(employeeAtt))
grep("YearsAtCompany", colnames(employeeAtt))
grep("OverTime", colnames(employeeAtt))
grep("MaritalStatus", colnames(employeeAtt))
grep("BusinessTravel", colnames(employeeAtt))
grep("StockOptionLevel", colnames(employeeAtt))
grep("WorkLifeBalance", colnames(employeeAtt))
grep("YearsSinceLastPromotion", colnames(employeeAtt))
````
Changing the variables
```{r}
#change the variables 
#change variables to factors and rename the levels to factored values
employeeAtt$BusinessTravel <- as.factor(employeeAtt$BusinessTravel)
employeeAtt$Department <- as.factor(employeeAtt$Department)
employeeAtt$EducationField <- as.factor(employeeAtt$EducationField)
employeeAtt$Gender <- as.factor(employeeAtt$Gender)
employeeAtt$JobRole <- as.factor(employeeAtt$JobRole)
employeeAtt$MaritalStatus <- as.factor(employeeAtt$MaritalStatus)
employeeAtt$OverTime <- as.factor(employeeAtt$OverTime)

levels(employeeAtt$Gender) <- c(1,2)
levels(employeeAtt$Gender)

#OverTime levels: 1=Yes 2=No
levels(employeeAtt$OverTime) <- c(1,2)
levels(employeeAtt$OverTime)

#MaritalStatus levels: 1=Divorced 2=Married 3=Single
levels(employeeAtt$MaritalStatus) <- c(1,2,3)
levels(employeeAtt$MaritalStatus)

#BusinessTravel levels:1=Non-Travel, 2=Travel Frequently 3=Travel Rarely
levels(employeeAtt$BusinessTravel) <- c(1,2,3)
levels(employeeAtt$BusinessTravel)

#Department levels: 1=Human Resources(HR) 2=Research & Development(R&D) 3= Sales
levels(employeeAtt$Department) <- c(1,2,3)
levels(employeeAtt$Department)

#EducationField levels: 1= Human Resources(HR) 2=Life Sciences 3=Marketing
# 4=Medical 5=Other 6=Technical Degree
levels(employeeAtt$EducationField) <- c(1,2,3,4,5,6)
levels(employeeAtt$EducationField)

#JobRole levels: 1=Healthcare Representative 2=Human Resources 3=Laboratory Tech
# 4=Manager 5=Manufacturing Director 6=Research Director
# 7=Research Scientist 8=Sales Executive 9=Sales Representative
levels(employeeAtt$JobRole) <- c(1,2,3,4,5,6,7,8,9)
levels(employeeAtt$JobRole)
````


#Building a KNN Model
```{r}
splitPerc = 0.7
  trainIndices2 = sample(seq(1: length(employeeAtt$ID)), round(splitPerc * length(employeeAtt$ID)))
  employeeTrain2 = employeeAtt[trainIndices2, ]
  employeeTest2 = employeeAtt[-trainIndices2, ]
  
  employeeTrain2= na.omit(employeeTrain2)
  employeeTest2 = na.omit(employeeTest2)

classifications = knn(employeeTrain2[,c("MonthlyIncome", "JobLevel", "TotalWorkingYears", "DistanceFromHome", "Age", "PerformanceRating", "YearsAtCompany", "YearsSinceLastPromotion"
                                      ,"OverTime", "MaritalStatus", "BusinessTravel", "StockOptionLevel", "WorkLifeBalance")], 
                      employeeTest2[,c("MonthlyIncome", "JobLevel", "TotalWorkingYears", "DistanceFromHome", "Age", "PerformanceRating", "YearsAtCompany", "YearsSinceLastPromotion"
                                      ,"OverTime", "MaritalStatus", "BusinessTravel", "StockOptionLevel", "WorkLifeBalance")], 
                      employeeTrain2$Attrition, prob = TRUE, k = 100)
table(classifications, employeeTest2$Attrition)
cm = confusionMatrix(table(classifications, employeeTest2$Attrition))


dim(employeeTrain2)
dim(employeeTest2)


table(classifications,employeeTest$Attrition)
confusionMatrix(table(classifications,employeeTest$Attrition))

#The Knn Model produce a specificty of 80% and a sensitivity of 61%. 

```


#Salary Income Model
````{r}
salLM <- lm(MonthlyIncome~JobLevel+TotalWorkingYears+YearsWithCurrManager,data=employee)
summary(salLM)

#The RMSE model is 1378 which is well below the 3000 threshold.

Salary_Prediction = predict(salLM,employeeSal)
Salary_Prediction
Salary_Prediction2<- data.frame(MonthlyIncome = Salary_Prediction)
Salary_Prediction = data.frame(ID = employeeSal[1], MonthlyIncome = Salary_Prediction2)

setwd("/Users/alrousan95/Desktop/SMU Data Science Masters/DS 6306")

write.csv(Salary_Prediction, "Salary_Prediction.csv", row.names=FALSE)
```

#Attrtion Prediction
````{r}
model = naiveBayes(Attrition~ MonthlyIncome +JobLevel+JobRole+TotalWorkingYears+DistanceFromHome+
                        Age+PerformanceRating+YearsAtCompany+YearsSinceLastPromotion+OverTime
                      + MaritalStatus + BusinessTravel + StockOptionLevel 
                      + WorkLifeBalance + YearsAtCompany, data = employeeTrain, laplace = 1)
employeeAtt$Attrition = "Attrition"
employeeAtt <- employeeAtt %>% relocate(Attrition, .before = Age)
attPred = predict(model,employeeAtt)
attPred
attPred2<- data.frame(Attrition = attPred)
attPred = data.frame(ID = employeeAtt[1], Attrition = attPred2)
levels(attPred$Attrition) <- c("No", "Yes")
levels(attPred$Attrition)

write.csv(attPred, "Attrition_Prediction.csv", row.names=FALSE)

````

```{r}
# Conclusion
 We were able to see that Attrition factored into Monthly Income, and the factors that come along with it are marital status, business travel, overtime and stock options.
The biggest factors for predicting Attrition were monthly income, business travel, and stock options. These along with 10 other factors were able to predict Attrition above 60% specificty and sensitivity.

Provided below are the links to the presentation and the rshiny app and are also provided at the end of the powerpoint.
````

